\documentclass[pdf]{beamer}
\mode<presentation>{}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{graphics,hufflen-macros}

\usetheme{Goettingen}

\addtobeamertemplate{footline}{\hfill\insertpagenumber/\insertframenumber/%
 \inserttotalframenumber\hspace*{0.2\textwidth}}

\newcommand{\AllTeXplus}{\texorpdfstring{\AllTeX}{(La)\TeX}}
\newcommand{\Dash}{---}

\def\LuaLaTeX{\textsf{Lua\LaTeX}}%

\DeclareRobustCommand{\Xe}{\leavevmode
 \tubhideheight{\hbox{X%
   \setbox0=\hbox{\TeX}\setbox1=\hbox{E}%                                      
   \lower\dp0\hbox{\raise\dp1\hbox{\kern\XekernbeforeE\tubreflect{E}}}%        
   \kern\XekernafterE}}}

\def\tubhideheight#1{\setbox0=\hbox{#1}\ht0=0pt \dp0=0pt \box0 }

\def\tubreflect#1{%
 \ifdim\fontdimen1\font>0pt%
   \raise1.75ex\hbox{\kern.1em\rotatebox{180}{#1}}\kern-.1em%
  \else\scalebox{-1}[1]{#1}%
 \fi}

\def\XekernafterE{-.1667em}
\def\XekernbeforeE{-.125em}
\def\XeLaTeX{\Xe{\kern.11em \LaTeX}}

\title{Extracting Information \\ from \AllTeXplus\ Source Files}
\author{Jean-Michel Hufflen}
\date{24~July 2022}
\institute{TUG 2022}

\begin{document}

\frame{\titlepage}

\section*{Contents}

\begin{frame}
\tableofcontents[]
\end{frame}

\section{Introduction}

\begin{frame}{\TeX\ldots}
is a wonderful tool for typesetting texts $\Longleftarrow$ well known;\pause

knows only its own formats $\Longleftarrow$ well known, too.\pause

Some information belonging to \AllTeX\ source files may be usable for other
purposes than typesetting, e.g., generating metadata for Web search
engines.\pause

\AllTeX's commands can do such jobs, but this is \emph{misuse} and complicates
the writing of classes. \TeX\ has not been designed for that, it is preferable
to use modern programming languages, with more suitable structures.
\end{frame}

\begin{frame}{Functional programming}
Functions are first-class objects, as other data.\pause

Functions can be results of a computation.\pause

So we can easily write \emph{generators} of functions.\pause

\pgScheme\ $\Longleftarrow$ elegant, data and programs have the same format, as
in any \pgname{Lisp} dialect.
\end{frame}

\section{Description}

\begin{frame}{Building a \emph{parsing function}}
\begin{center}
\texttt{(g-mk-tex-parsing-f \textsl{directive} \ldots)}
\end{center}
All the \emph{directives} are grouped, `compiled' into a function ready to
parse a source file.
\end{frame}

\begin{frame}{Directives}
\begin{small}
\begin{ttfamily}
\begin{tabbing}
(g-retain-command \= \textsl{command-name} \textsl{arg-nb} 
\textsl{optional-arg?} \\
 \> \textsl{top-level?}\ \textsl{recursive?}\ \textsl{preamble?} \\
 \> \textsl{occ-nb-info} \textsl{function})  
\end{tabbing}
\end{ttfamily}
where:
\begin{description}
 \item[\textnormal{\texttt{\textsl{command-name}}}] is the name of the command
to be caught;
 \item[\textnormal{\texttt{\textsl{arg-nb}}}] is the argument number of this
command;
 \item[\textnormal{\texttt{\textsl{optional-arg?}}}] is true if the first
argument is optional, surrounded by square brackets, false otherwise;
 \item[\textnormal{\texttt{\textsl{top-level?}}}] is true if this command is to
be searched only at the top level, false otherwise;
 \item[\textnormal{\texttt{\textsl{recursive?}}}] is used when \command{input}
commands are encountered: if it is true, corresponding files are searched
recursively, otherwise such an \command{input} command is just skipped;
 \item[\textnormal{\texttt{\textsl{preamble?}}}] stops searching after a
preamble if it is bound to true; otherwise, goes on.
\end{description}

\end{small}
\end{frame}

\begin{frame}{Other arguments}
\begin{description}
 \item[\textnormal{\texttt{\textsl{occ-nb-info}}}] may be bound to:
 \begin{itemize}
  \item \texttt{0} or the false value (\texttt{\#f}): the command should not
appear within the file, this is checked;
  \item a positive integer $n$: the first $n$ occurrences of this command are
processed, and following ones are ignored;
  \item the true value (\texttt{\#t}): all the occurrences of this command are
processed;
 \end{itemize}
 \item[\textnormal{\texttt{\textsl{function}}}] the \pgScheme\ function to
call, it \emph{must} have the same number of arguments than
\command{\textsl{command-name}}.
\end{description}
\end{frame}

\begin{frame}{Directives \emph{(con'd)}}
\begin{ttfamily}
\begin{tabbing}
(g-retain-match \= \textsl{command-name} \textsl{s} \textsl{top-level?}\ \\
 \> \textsl{recursive?}\ \textsl{preamble}\ \textsl{occ-nb-info} \\
 \> \textsl{function})
\end{tabbing}
\end{ttfamily}
is used when \command{\textsl{command-name}}'s arguments are expressed by means
of a pattern, e.g., \texttt{"\#1\textbackslash endcsname"} for the
\command{csname} command. \texttt{\textsl{s}} is a string bound to such a
pattern, the other arguments have the same meaning than
\texttt{g-retain-command}'s.\pause

The arguments of corresponding functions in \pgScheme\ are strings in both
cases.
\end{frame}

\begin{frame}{Result's result}
\texttt{g-mk-tex-parsing-f} builds a function that applies to a filename and
returns:
\begin{description}
 \item[\textnormal{\texttt{false}}] if something went wrong, or a forbidden
command is included into the file;
 \item[\textnormal{\texttt{true}}] in all other cases. 
\end{description}\pause

You have to update your own structures when a file is parsed. If an error
occurs, they may be in an inconsistent state.
\end{frame}

\begin{frame}{Destructuring}
These functions can be used to destructure an optional
argument\Dash\texttt{\textsl{s},\textsl{s}$_{0}$} are strings\Dash:\pause
\begin{description}
 \item[\textnormal{\texttt{(g-parse-to-list \textsl{s})}}] returns the elements
of a comma-separated list;
 \item[\textnormal{\texttt{(g-parse-to-alist \textsl{s} \textsl{s}$_{0}$)}}]
returns the successive associations of a comma-separated list whose elements 
are \texttt{\textsl{key=value}} pairs; if a key is given without a value, this
missing value is replaced by \texttt{\textsl{s}$_{0}$}.
\end{description}
In both cases, the original order is preserved.\pause

\textbf{Remark} \quad Note that \texttt{g-parse-to-list},
\texttt{g-parse-to-alist}, \texttt{g-retain-command} and
\texttt{g-retain-match} are functions, whereas \texttt{g-mk-tex-parsing-f} is a
\emph{macro}.
\end{frame}

\begin{frame}{Example}
Considering a source text for \LaTeX, give:
\begin{itemize}
 \item the used options of the \packagename{babel} package,
 \item the title,
 \item the number of occurrences of the \command{emph} command.
\end{itemize}

(Show.)
\end{frame}

\section{Discussion}

\begin{frame}{`History'}
When I realised \mlBibTeX, I needed to know which options of the
\packagename{babel} package were used. Only the preamble was to be
searched.\pause

The same to determine which encoding was used.\pause

One year and a half ago, I became the new editor of the journal of the
French-speaking \TeX\ user group.\pause

I decided to revise the class used for this journal and discovered that the
previous class was used to build other files, such as metadata for Web search
engines.\pause

From my point of view, \TeX\ is intended for typesetting, and other tasks
should be delegated to more modern programming languages.\pause

I could have used \pgname{Lua}\ldots\pause\ but \pgname{Lua\TeX} was unable to
process some texts designed for pdf\TeX\ or \XeLaTeX.
\end{frame}

\begin{frame}{Principles}
Simple extractions\Dash e.g., a title\Dash must be simple.\pause

If addressing occurrences of a \TeX\ command inside a source text, it should be
possible to catch it.\pause

But more experience will be needed.
\end{frame}

\section{Presently}

\begin{frame}{Current state}
Some bugs to fix for the pattern-matching of \TeX\ command's arguments.\pause

The other points seems to be OK.\pause

Available as a \pgScheme\ library.
\end{frame}

\section{Conclusion}

\begin{frame}{Ending}
For many years, we have seen that in addition to \TeX's works, many other tasks
are more related to `classical' programming. In particular, that is why
\pgname{Lua\TeX} emerged.\pause

We relate our work to avoiding \emph{information redundancy}.
\end{frame}

\end{document}
